--- 
title: View Gene Body First Change Point
author: Robert Shear
output:
  pdf_document: default
params:
  display.window: 700
  minimum.fcp.location: 20
  experiement_filename: "/n/groups/churchman/rds19/data/S005/HresSE_1024_by_6.rds"
  sample.name: ["wt-1", "wt-2"]
  lower.trim.parameter: 0.15 # lower 'trim' cutoff, mean occupancy / nt
  running.mean.window: 5  # running mean width for smoothing img heat map
---
```{r setup, include=FALSE}

# TODO REVISE TO USE Final form of change points

knitr::opts_chunk$set(echo = FALSE, 
    warning = FALSE, 
    message = FALSE,
    fig.width = 6,
    fig.height = 4,
    dpi = 300)
sample.name <- params$sample.name
display.window <- params$display.window
minimum.fcp.location <- params$minimum.fcp.location
set.seed(6272017)

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(SummarizedExperiment)
  library(tidyverse)
})

# TODO Move to common utility script
# unroll an assay matrix in the order required by 
assay_as_vector <- function(exp, assayname) c(t(assay(exp, assayname)))
assay_as_list <- function(exp, assayname) lapply(c(t(assay(exp, assayname))), list)


```


This is an analysis of first change points (FCP) for sample **`r sample.name`**. We only consider the first **`r display.window`** bases. FCP's less than **`r minimum.fcp.location`** bases from either the 3' or 5' end are disregarded.


```{r}

e <- readRDS(params$experiement_filename)
e <- e[,params$sample.name]
assay(e, "mu") %>%
  as_tibble(m, rownames = "gene") %>%
  mutate(strand = as.character(strand(rowRanges(e)))) %>%
  pivot_longer(!c(gene, strand), names_to = "sample", values_to = "mean")  %>%
  mutate(k = assay_as_vector(e, "k"), 
         cpt = assay_as_list(e, "cpt"),
         segments = assay_as_vector(e, "segments"),
         seg_max_var = map_dbl(segments, function(u) max(u$v))
  ) %>%
  group_by(sample) %>% 
  mutate(quantile = cut(ecdf(mean)(mean), c(0,.25,.75,1), 
                        labels = c("Q1", "Q23", "Q4"))) %>%
  ungroup() -> x
```


Number of ORF's in sample: `nrow(e`.

TBD
Trimmed lower and upper 5%. Occupancy range of processed ORF's: 
`trim.range`

```{r}
# cds_ext is the GRanges object for the selected genes
cds_ext <- segments$cds_ext[gene.names]

cds_segment <- segments$cds_segment[as.character(segments$cds_segment$Parent) %in% gene.names]
cds_segment <- split(cds_segment, cds_segment$seq)
for (u in 1:2) {
  names(cds_segment[u][[1]]) <- cds_segment[u][[1]]$Parent
}

# TODO: there is a gap between seg1 end and seg2 start created in GenCliffs.R. Rivisit this
# extract breakpoints and sort ascending
idx <- width(cds_segment[1][[1]])
names(idx) <- cds_segment[1][[1]]$Parent

# Trim the windows...20 bases on the left and right
display.border <- 20
idx <- idx[idx >= display.border & idx <= display.window - display.border]

idx <- sort(idx)

img <- sapply(cds_ext[names(idx)], function(u) {
  s <- as.character(strand(u))
  w <- Rle(
    pn[s][[1]][as.character(seqnames(u))][[1]][start(u):end(u)])
  if (s == "-") {
    w <- rev(w)
  }
  w <- as.numeric(w)
  length(w) <- display.window
  # normalize by gene
  w / sum(w, na.rm = TRUE) * width(u)
})

color.map <- colorRampPalette(c("yellow", "orange", "red", "brown"), space = "Lab")(30)
```

RNAP II log_2 occupancy, normalized by gene, unit occupancy per nt, sorted by position of first change point.

```{r}
# display relative density as heat map
x <- log2(img)
x[which(is.na(x) | x <= -100)] <- min(x[x > -100], na.rm = TRUE)
r <- dim(x)
rng <- range(x)
{
  image.plot(x = 1:r[2], y = 1:r[1],
    z = t(x),
    zlim = rng,
    col = color.map,
    ylab = "NT from TSS", xlab = "gene", 
    legend.lab = "normalized log2 occupancy"
  )
#  lines(idx, 1:length(idx), col = "black")
}
```

RNAPII Occupancy, sorted by position of first change point. Each gene is visually stretched or compressed so that its FCP is in the center of the graph.

```{r}
# stretch / shrink the tracks so that the CP is in the middle
# mapping image to stretch/shrink


# TOD: Use maatrix traansformations
img.xformed <- sapply(seq_along(idx), function(gene.id) {
  cp <- idx[gene.id]
  # stretch the first segment
  z <- round(((0:(cp - 1) * display.window / 2 / cp)),digits = 0)
  seg1 <- rep(img[1:cp, gene.id], c(z[2:length(z)], round(display.window / 2,0)) - z)
  # grab the second segment
  seg2.data <- img[(cp + 1):display.window, gene.id]
  seg2.data <- seg2.data[!is.na(seg2.data)]
  # now shrink it
  seg2.idx <- findInterval(seq(length(seg2.data)) * display.window / 2 / length(seg2.data), seq(display.window / 2))
  seg2 <- sapply(1:(display.window / 2), function(u) mean(seg2.data[seg2.idx == u]))
  c(seg1, seg2)
})

# TODO: Occpancy vs. Normalized Occupancy vs. Occupancy adjusted for ORF length?
{
  image.plot(x = ((1:nrow(img.xformed)) - nrow(img.xformed) / 2) / nrow(img.xformed) * 2, 
      y = 1:ncol(img.xformed), z
      = log2(img.xformed), 
      zlim = rng,
      col = color.map,
      xlab = "Change point shifted to middle", 
      ylab = "gene", legend.lab = "normalized log2 occupancy")
  abline(v = 0, col = "black")
}
```

Attempt linear regression of occupancy vs. nucleotide positon on first and segment. How homogeneous are the segments?

```{r}
# what do the slopes of the first segment look like?
slope.first <- sapply(idx, function(i) {
  lo <- 1
  hi <- idx[i] - 1
  coef(glm(x ~ t, data = data.frame(t = lo:hi, x = img[lo:hi,i])))["t"]
  }
)

slope.next100 <- sapply(idx, function(i) {
  lo <- idx[i] + 1
  hi <- idx[i] + 100
  coef(glm(x ~ t, data = data.frame(t = lo:hi, x = img[lo:hi, i])))["t"]
  }
)

boxplot(slope.first, slope.next100, names = c("1:FCP-1","FCP+1:FCP+100"),
        main = paste("Intrasegment slopes. Sample", sample.name), 
        sub = "Slope of linear reg within segment", ylab = "slope",
        outline = FALSE, notch = TRUE)
abline(h = 0, col = "blue")
```

The metagene.

```{r}
meta.cp.min <- 100
x <- sapply(which(idx > meta.cp.min & idx + meta.cp.min < display.window), function(u) {
  cp <- idx[u]
  img[(cp - meta.cp.min):(cp + meta.cp.min), u]
})
plot(seq(from = -meta.cp.min, to = meta.cp.min), rowMeans(x, na.rm = TRUE), type = 'l',
     main = paste("Mean occupancy aligned to first change point", sample.name),
     xlab = "NT from FCP",
     ylab = "Mean occupancy")
```

Is the location of the FCP correlated with occupancy?


```{r}

{
  plot(log2(mcols(cds_ext[names(idx)])$cliff), log2(mcols(cds_ext[names(idx)])$mean), 
     main = "Cliff direction",
     xlab = "log2 occupancy change at cliff",
     ylab = "log2 mean occupancy")
  abline(v = 0, col = "red")
}

cliff_direction <- table(sign(log(cds_ext$cliff)))
step_down <- cliff_direction["1"]
pct_step_down <- step_down / sum(cliff_direction)

```

For this sample, what percentage of the FCP's are "step-down" vs "step-up"?


N=`r sum(cliff_direction)`, step down=`r step_down` (`r round(pct_step_down*100,1)`%)



